Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-02 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Our motivation for estimating transient strain rates is, in part, to
detect geophysical phenomena. As we will see, geophysical signal can
be easily identified by visually inspecting the solution for $\strain$
from eqs. (\ref{eq:StrainMean}) and (\ref{eq:StrainCov}). However, if
we want to detect geophysical phenomena automatically, then we need to
define a detection criterion. We use a signal-to-noise ratio, SNR,
that is based on the Frobenius norm of $\strain$,
$||\strain||_\mathrm{F} = \left(\strain_{\e\e}^2 + \strain_{\n\n}^2 +
2\strain_{\e\n}^2\right)^{\frac{1}{2}}$, for our detection criterion.
In the geodetic literature, $||\dot{\varepsilon}||_\mathrm{F}$ is
often used as a metric for the strain rate ``magnitude", and it is
sometimes referred to as the second invariant of strain rate. Noting
that $||\strain||_\mathrm{F}$ is a random variable, we take SNR to be
the ratio of the estimated mean and standard deviation of
$||\strain||_\mathrm{F}$. An estimate of the mean is found by
evaluating $||\strain||_\mathrm{F}$ at the mean of $\strain$,
\begin{align}
\mu_{||\strain||_\mathrm{F}} &\approx ||\strain||_F \big|_{\strain = \mu_{\strain}}\nonumber \\
                             &= \left(\mu_{\strain_{\e\e}}^2 + 
                                      \mu_{\strain_{\n\n}}^2 + 
                                      2\mu_{\strain_{\e\n}}^2 \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard
deviation,
\begin{equation}
\sigma_{||\strain||_\mathrm{F}} \approx
\left( \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\e}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\e}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\n\n}}\right|_{\strain = \mu_{\strain}} \right)^2
       \sigma^2_{\strain_{\n\n}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\n}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\n}} \right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\strain_{ij}}(p) = C_{\strain_{ij}}(p,p)$. After some
calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\strain||_\mathrm{F}}(p)}{\sigma_{||\strain||_\mathrm{F}}(p)} \\
                &= \frac{\mu_{\strain_{\e\e}}(p)^2 +
                         \mu_{\strain_{\n\n}}(p)^2 +
                         2\mu_{\strain_{\e\n}}(p)^2}
                        {\big(\sigma^2_{\strain_{\e\e}}(p)\mu_{\strain_{\e\e}}(p)^2 + 
                              \sigma^2_{\strain_{\n\n}}(p)\mu_{\strain_{\n\n}}(p)^2 + 
                              4\sigma^2_{\strain_{\e\n}}(p)\mu_{\strain_{\e\n}}(p)^2
                         \big)^{\frac{1}{2}}}.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it
identifies the position and time of anomalous deformation. We can
reasonably suspect that some transient geophysical phenomena is
occurring wherever and whenever SNR is larger than ${\sim}3$.

@@

We next establish our prior model for transient displacements.
Specifically, we discuss our choice for the spatial and temporal
functions making up $C_{u_i}$, $X_i$ and $T_i$. For $X_i$, we use the
squared exponential (SE) covariance function,
\begin{equation}\label{eq:SE}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2 \ell^2}\right).
\end{equation}
The SE covariance function is commonly used for kriging
\citep[e.g,][]{Cressie1992} and Gaussian process regression
\citep[e.g.,][]{Rasmussen2006}. In terms of geodetic applications,
\citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE
covariance function accurately describes the spatial covariance of
secular GNSS derived velocities in Japan.

@@

Having established a noise model and a prior for transient
displacements, we can now use the GNSS data to calculate transient
strain rates, $\strain$, in the Puget Sound region. We evaluate
$\strain$ at a grid of points spanning the study area for each day
from January 1, 2010 to May 15, 2017. In Figure \ref{fig:StrainMap},
we show a map view of $\strain$ on January 1, 2016, which coincides
with the height of the winter 2015-2016 SSE. In Figure
\ref{fig:StrainTs}, we show a time series of $\strain$ at a position
on the Olympic Peninsula, which tends to be where $\strain$ is the
largest during SSEs. We also include a supplementary animation showing
a map view of $\strain$ over time.

@@

The strain rates shown in Figure \ref{fig:StrainMap} are generally
similar to the strain rates during the other six SSEs considered in
this study. The SSEs cause trench perpendicular compression in the
Olympic Peninsula and extension east of Puget Sound. The strain
transitions from compression to extension around the southern tip of
Vancouver Island, which coincides with the location of thrust slip for
SSEs in the Puget Sound region \citep[e.g.,][]{Dragert2001, Wech2009,
Schmidt2010}. Thus, this pattern of strain is to be expected. During
the period in between SSEs, secular strain rates indicate trench
perpendicular compression throughout this study region
\citep{Murray2000, McCaffrey2007, McCaffrey2013}. When comparing
inferred strain rates from SSEs to the secular strain rates, we see
that SSEs are concentrating tectonically accumulated strain energy
towards the trench, and presumably pushing the subduction zone closer
to failure.

%A key difference between the strain inferred here and strain that can
%be derived from fault slip models is that our estimated strain rates
%are not based on an assumed physical model. In contrast, fault slip
%models can be biased by errors in the assumed fault geometry or
%lithospheric rheology. Moreover, the degrees of freedom in fault slip
%models usually cannot be constrained by GNSS data alone, and it is
%necessary to impose regularization which further biases the results.
%Since our estimated strain rates lack such systematic errors, we can
%be more confident that our solution is unbiased and has meaningful
%uncertainties.

@@

To verify that the estimated transient strain rates are accurately
identifying geophysical signal, we compare the signal-to-noise ratio
from eq. (\ref{eq:SNR}) to the frequency of seismic tremor (Figure
\ref{fig:StrainMag}). A signal-to-noise ratio greater than ${\sim}3$
can be interpreted as a detected geophysical signal. We detect nine
distinct events, which each correspond to peaks in seismic tremor. The
smaller events detected in August 2014 and February 2017 can be
considered inter-SSE events. They were not among the SSEs used to
constrain the prior covariance function. In between peaks in seismic
tremor, the signal-to-noise ratio is consistently between 0 and 2,
suggesting that all the transient strain detected at this location is
associated with SSEs and inter-SSE events.

@@

The results we have presented thus far indicate that we are
identifying the strain that we should expect to see. There are,
however, subtle features in our estimated transient strain rates which
we were not expecting. For example, some SSEs are preceded by a brief
period of east-west extension on the Olympic Peninsula. This feature
can be seen several days before the summer 2012 and winter 2015-2016
SSEs in Figure \ref{fig:StrainTs} as well as in the supplementary
animation. While this deformation is noteworthy, a discussion on the
mechanisms causing it is outside the scope of this study.

>>>>

Our motivation for estimating transient strain rates is, in part, to
detect transient geophysical processes. As we will see, geophysical
signal can be easily identified by visually inspecting the solution
for $\strain$ from eqs. (\ref{eq:StrainMean}) and
(\ref{eq:StrainCov}). However, if we want to detect geophysical signal
automatically, then we need to define a detection criterion. We use a
signal-to-noise ratio, SNR, that is based on the Frobenius norm of
$\strain$, $||\strain||_\mathrm{F} = \left(\strain_{\e\e}^2 +
\strain_{\n\n}^2 + 2\strain_{\e\n}^2\right)^{\frac{1}{2}}$, for our
detection criterion. In the geodetic literature,
$||\dot{\varepsilon}||_\mathrm{F}$ is often used as a metric for the
strain rate ``magnitude", and it is sometimes referred to as the
second invariant of strain rate. Noting that $||\strain||_\mathrm{F}$
is a random variable, we take SNR to be the ratio of the estimated
mean and standard deviation of $||\strain||_\mathrm{F}$. An estimate
of the mean is found by evaluating $||\strain||_\mathrm{F}$ at the
mean of $\strain$,
\begin{align}
\mu_{||\strain||_\mathrm{F}} &\approx ||\strain||_F \big|_{\strain = \mu_{\strain}}\nonumber \\
                             &= \left(\mu_{\strain_{\e\e}}^2 + 
                                      \mu_{\strain_{\n\n}}^2 + 
                                      2\mu_{\strain_{\e\n}}^2 \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard
deviation,
\begin{equation}
\sigma_{||\strain||_\mathrm{F}} \approx
\left( \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\e}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\e}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\n\n}}\right|_{\strain = \mu_{\strain}} \right)^2
       \sigma^2_{\strain_{\n\n}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\n}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\n}} \right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\strain_{ij}}(p) = C_{\strain_{ij}}(p,p)$. After some
calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\strain||_\mathrm{F}}(p)}{\sigma_{||\strain||_\mathrm{F}}(p)} \\
                &= \frac{\mu_{\strain_{\e\e}}(p)^2 +
                         \mu_{\strain_{\n\n}}(p)^2 +
                         2\mu_{\strain_{\e\n}}(p)^2}
                        {\big(\sigma^2_{\strain_{\e\e}}(p)\mu_{\strain_{\e\e}}(p)^2 + 
                              \sigma^2_{\strain_{\n\n}}(p)\mu_{\strain_{\n\n}}(p)^2 + 
                              4\sigma^2_{\strain_{\e\n}}(p)\mu_{\strain_{\e\n}}(p)^2
                         \big)^{\frac{1}{2}}}.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it
identifies the position and time of anomalous deformation. We can
reasonably suspect that some transient geophysical phenomena is
occurring wherever and whenever SNR is larger than ${\sim}3$.

@@

We next establish a prior model for transient displacements.
Specifically, we discuss our choice for the spatial and temporal
functions making up $C_{u_i}$, $X_i$ and $T_i$. For $X_i$, we use the
squared exponential (SE) covariance function,
\begin{equation}\label{eq:SE}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2 \ell^2}\right).
\end{equation}
The SE covariance function is commonly used for kriging
\citep[e.g,][]{Cressie1992} and Gaussian process regression
\citep[e.g.,][]{Rasmussen2006}. In terms of geodetic applications,
\citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE
covariance function accurately describes the spatial covariance of
secular GNSS derived velocities in Japan.

@@

Having established a noise model and a prior for transient
displacements, we can now calculate transient strain rates, $\strain$,
in the Puget Sound region from the GNSS data. We evaluate $\strain$ at
a grid of points spanning the study area for each day from January 1,
2010 to May 15, 2017. In Figure \ref{fig:StrainMap}, we show a map
view of $\strain$ on January 1, 2016, which coincides with the height
of the winter 2015-2016 SSE. In Figure \ref{fig:StrainTs}, we show a
time series of $\strain$ at a position on the Olympic Peninsula, which
is where $\strain$ tends to be the largest during SSEs. We also
include a supplementary animation showing a map view of $\strain$ over
time. In each figure, we show the mean and standard deviation of
$\strain$, making it easy to identify which features are statistically
significant.

@@

The transient strain rates shown for the winter 2015-2016 SSE in
Figure \ref{fig:StrainMap} are characteristic of most SSEs in the
Puget Sound region. The SSEs cause trench perpendicular compression in
the Olympic Peninsula and extension east of Puget Sound. The strain
transitions from compression to extension around the southern tip of
Vancouver Island, coinciding with where fault slip tends to be
inferred \citep[e.g.,][]{Dragert2001, Wech2009, Schmidt2010}. Thus,
this pattern of strain is to be expected. Over decadal timescales,
there is trench perpendicular compression throughout this study region
caused by steady tectonic plate motion \citep{Murray2000,
McCaffrey2007, McCaffrey2013}. When comparing the transient strain
rates caused by SSEs to the secular strain rates, we see that SSEs are
concentrating tectonically accumulated strain energy towards the
trench, and presumably pushing the subduction zone closer to failure.

@@

To verify that the estimated transient strain rates are accurately
identifying geophysical signal, we compare the signal-to-noise ratio
from eq. (\ref{eq:SNR}) at a point on the Olympic Peninsula to the
frequency of seismic tremor (Figure \ref{fig:StrainMag}). A
signal-to-noise ratio greater than ${\sim}3$ can be interpreted as a
detected geophysical signal. We detect nine distinct events, which
each correspond to peaks in seismic tremor. The smaller events
detected in August 2014 and February 2017 can be considered inter-SSE
events. They were not among the SSEs used to constrain the prior
covariance function. In between peaks in seismic tremor, the
signal-to-noise ratio is consistently between 0 and 2, suggesting that
all the transient strain detected at this point on the Olympic
Peninsula is associated with SSEs and inter-SSE events.

@@

% note that we are only discussing strain with known origin

The results we have presented indicate that we are identifying the
transient strain that we should expect to see. This is not to say that
there are no unexpected features in $\strain$ that are worth further
exploration. However, a discussion on features in $\strain$ that have
unknown origin would be outside the scope of this paper.
